---
title: "Thresholds"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi = 200, fig.width = 8)

library(here)
library(lubridate)
library(plotly)
library(stringr)
library(summaryplots)

source(here("pages/import_data_and_params.R"))

# filter depth data
dat_depth <- readRDS(here("pages/data/depth_rolling_sd_reprocessed.rds")) %>%
  select(
    -c(int_sample, n_sample, rolling_sd_flag_sensor_depth_measured_m)
  ) %>%
  ss_pivot_longer() %>%
  filter(
    # sensor read outside range
    !(station == "Olding Island" &
        deployment_range == "2022-Jun-03 to 2022-Sep-30" &
        depth_log == 3),
    # freshwater stations
    !(station %in% c("Piper Lake", "Hourglass Lake", "0193", "Sissiboo")),
    # suspect range
    !(station == "Long Beach" &
        deployment_range == "2020-Jul-16 to 2020-Nov-29" & depth_log == 5),
    !(station == "Sandy Cove" &
        deployment_range == "2020-Jul-16 to 2020-Nov-30" & depth_log == 5),
    !(station == "Tickle Island 1" &
        deployment_range == "2020-Oct-21 to 2021-Aug-25" & depth_log == 5)
  ) %>% 
  mutate(month = month(timestamp_utc, label = TRUE)) %>% 
  ss_pivot_wider()
```


QC test thresholds for measured depth were based on historical Coastal Monitoring Program data submitted to the Nova Scotia Open Data Portal in December 2022. Preliminary QC was applied to the historical data. Obvious outliers and freshwater stations were omitted.

## Gross Range Test

### Sensor Thresholds

The sensor thresholds were determined based on the aquaMeasure and VR2AR manufacturer specifications (@tbl-depth-sensor-thresh).

```{r}
#| label: tbl-depth-sensor-thresh
#| tbl-cap: Depth sensor thresholds for the Gross Range Test.

sensors %>%
  filter(str_detect(variable, "depth")) %>%
  mutate(
    `Sensor (link to spec sheet)` =
      paste0('<a  target=_blank href=', url, '>', sensor_type,'</a>')
  ) %>%
  select(`Sensor (link to spec sheet)`, sensor_min, sensor_max) %>%
  distinct() %>% 
  datatable(options = dt_options2, rownames = FALSE, escape = FALSE)
```


### User Thresholds

Measured sensor depth is a function of where the sensor is located on the string and is not necessarily related to the historical measurements (@fig-depth-hist). For example, sensors in Antigonish County have typically been moored in relatively shallow depths, but deeper sensors could be deployed there in the future.

To avoid flagging future observations at unusual depths, a single set of user thresholds was manually selected and applied to the pooled data (0–120 m; @tbl-depth-user).

:::panel-tabset

#### Figure 1
```{r}
#| warning: false
#| fig-height: 10
#| label: fig-depth-hist
#| fig-cap: Distribution of depth observations (binwidth = 2 m). Dotted orange lines indicate the user thresholds.
#| fig-alt: Histogram of measured depth observations.

# user_depth <- dat_depth %>% 
#   group_by(county) %>% 
#   qc_calculate_user_thresholds(
#     var = "sensor_depth_measured_m",
#     n_sd = 3, keep_stats = TRUE
#   ) %>% 
#   pivot_wider(names_from = threshold, values_from = threshold_value) %>% 
#   select(county, mean = mean_var, stdev = sd_var, user_min, user_max) 

p <- plot_histogram(
  dat_depth, hist_col = sensor_depth_measured_m, 
  binwidth = 2,
  pal= hist_pal
) +
  scale_x_continuous("Measured Sensor Depth (m)") +
  facet_wrap(~county, ncol = 2) +
  geom_vline(
    xintercept = c(0, 120), col = "#EDA247", linetype = 3) +
  theme_facet_plotly

ggplotly(p, tooltip = "text")
```

#### User Thresholds

```{r}
#| label: tbl-depth-user
#| tbl-cap: Gross Range Test statistics and user thresholds for measured depth.

thresholds %>% 
  filter(variable == "sensor_depth_measured_m", qc_test == "grossrange") %>% 
  pivot_wider(names_from = threshold, values_from = threshold_value) %>% 
  select(user_min, user_max) %>% 
  na.omit() %>% 
  datatable(options = dt_options2, rownames = FALSE)
```
:::

## Climatological Test

The Climatological Test was not applied to the measured depth data because depth has no expected or observed seasonal cycle (@fig-depth-clim; @fig-depth-hist-clim).

::: panel-tabset

#### Figure 2
```{r}
#| fig-height: 10
#| label: fig-depth-clim
#| message: false
#| fig-cap: Mean ± 3 standard deviations of the monthly depth observations.
#| fig-alt: Mean monthly measured depth shown as points, with ± 3 standard deviations as vertical lines.

depth_clim <- dat_depth %>% 
  group_by(county, month) %>% 
  summarise(
    mean = mean(sensor_depth_measured_m), 
    stdev = sd(sensor_depth_measured_m)
  ) %>% 
  mutate(
    variable = "sensor_depth_measured_m",
    season_min = round(mean - 3 * stdev, digits = 1),
    season_max = round(mean + 3 * stdev, digits = 1)
  )

p <- plot_mean_sd_season(depth_clim, n_sd = 3, facet_county = FALSE) +
  scale_y_continuous("Measured Sensor Depth (m)") +
  facet_wrap(~county, ncol = 2) +
  theme_facet_plotly
  
ggplotly(p)

```

#### Figure 3
```{r}
#| warning: false
#| fig-height: 10
#| label: fig-depth-hist-clim
#| message: false
#| fig-cap: Monthly distribution of measured depth observations (binwidth = 2 m).
#| fig-alt: A separate panel for each month showing a histogram of measured depth in that month.

p <- dat_depth %>%
  plot_histogram(
    hist_col = sensor_depth_measured_m, binwidth = 2,
    pal= hist_pal
  ) +
  scale_x_continuous("Measured Sensor Depth (m)") +
  facet_wrap(~month,  ncol = 3) +
  theme_facet_plotly

ggplotly(p, tooltip = "text")
```

:::

## Spike Test

Separate thresholds were calculated for each county and sensor type to account for the substantial spatial differences in tidal range around the province and the difference in precision between the types of sensors that measure depth (@fig-depth-spike-dist-a; @fig-depth-spike-dist-b).

For all groups, the distribution of the [spike value](qc_tests.qmd#sec-spike-test) is skewed right, and so several upper percentile values (90th, 95th, and 99.7th percentile) were evaluated to use as the $\text{spike}_\text{low}$. There were relatively few large single-value spikes in the depth data, and so the 99.7th percentile was selected to avoid false positives. $\text{spike}_\text{high}$ was defined as $3 \times \text{spike}_\text{low}$ to identify especially large spike values.

In Digby and Pictou counties, the only depth observations have been from VR2AR sensors. In the future, depth may be measured with aquaMeasure sensors in these counties. If this happens, the VR2AR thresholds will be used for the Spike Test until the threshold calculation exercise is repeated.

::: panel-tabset

### Figure 4A
```{r}
#| warning: false
#| fig-height: 10
#| label: fig-depth-spike-dist-a
#| fig-cap: Distribution of the spike value of depth observations measured by aquaMeasure sensors (binwidth = 0.1 m). Dotted orange line indicates $\text{spike}_\text{low}$; dotted red line indicates $\text{spike}_\text{high}$. Note difference in y-axis scale.
#| fig-alt: Histogram of the spike value calculated for depth observations measured by aquaMeasure sensors.

spike_thresh <- thresholds %>% 
  filter(variable == "sensor_depth_measured_m", qc_test == "spike") %>% 
  pivot_wider(names_from = threshold, values_from = threshold_value) 

dat_sp <- dat_depth %>% 
  group_by(county, station, deployment_range, sensor_serial_number) %>%
  dplyr::arrange(timestamp_utc, .by_group = TRUE) %>%
  mutate(
    lag_value = lag(sensor_depth_measured_m),
    lead_value = lead(sensor_depth_measured_m),
    spike_ref = (lag_value + lead_value) / 2,
    spike_value = abs(sensor_depth_measured_m - spike_ref)  
  ) %>% 
  ungroup() %>% 
  filter(!is.na(spike_value))

p <- dat_sp %>% 
  filter(sensor_type == "aquameasure", !is.na(spike_value)) %>%
  plot_histogram(hist_col = spike_value, binwidth = 0.1, pal = hist_pal) +
  facet_wrap(~county, ncol = 2, scales = "free_y") +
  geom_vline(
    dat = filter(spike_thresh, sensor_type == "aquameasure"),
    aes(xintercept = spike_low), 
    col = "#EDA247", linetype = 3
  ) +
  geom_vline(
    dat = filter(spike_thresh, sensor_type == "aquameasure"), 
    aes(xintercept = spike_high), 
    col = "#DB4325", linetype = 3
  ) +
  scale_x_continuous("Measured Sensor Depth (m)\nSpike Value") +
  theme_facet_plotly2

ggplotly(p, tooltip = "text")

```

### Figure 4B
```{r}
#| warning: false
#| fig-height: 10
#| label: fig-depth-spike-dist-b
#| fig-cap: Distribution of the spike value of depth observations measured by VR2AR sensors (binwidth = 0.1 m). Dotted orange line indicates $\text{spike}_\text{low}$; dotted red line indicates $\text{spike}_\text{high}$. Note difference in y-axis scale.
#| fig-alt: Histogram of the spike value calculated for depth observations measured by VR2AR sensors.

p <- dat_sp %>% 
  filter(sensor_type == "vr2ar") %>% 
  plot_histogram(hist_col = spike_value, binwidth = 0.1, pal = hist_pal) +
  facet_wrap(~county, ncol = 2, scales = "free_y") +
  geom_vline(
    dat = filter(spike_thresh, sensor_type == "vr2ar"),
    aes(xintercept = spike_low), 
    col = "#EDA247", linetype = 3
  ) +
  geom_vline(
    dat = filter(spike_thresh, sensor_type == "vr2ar"), 
    aes(xintercept = spike_high), 
    col = "#DB4325", linetype = 3
  ) +
  scale_x_continuous("Measured Sensor Depth (m)\nSpike Value") +
  theme_facet_plotly2

ggplotly(p, tooltip = "text")

```

### Spike Thresholds

```{r}
#| label: tbl-depth-spike
#| tbl-cap: Spike thresholds for measured depth.

spike_thresh %>% 
  select(county, sensor_type, spike_low, spike_high) %>% 
  datatable(options = dt_options, rownames = FALSE)
```

:::

## Rolling Standard Deviation Test

The Rolling Standard Deviation Test thresholds were calculated and applied separately for each county due to spatial differences in tidal range.

The distribution of [rolling standard deviation](qc_tests.qmd#sec-rolling-sd-test) is approximately normal for each county (@fig-depth-roll-sd-dist), and so the mean and standard deviation were used to calculate $\text{rolling\_sd}_\text{max}$.

$\text{rolling\_sd}_\text{max} = \text{rolling\_sd}_\text{mean} + 3 \times \text{rolling\_sd}_\text{sd}$

::: panel-tabset

### Figure 5
```{r}
#| warning: false
#| fig-height: 10
#| label: fig-depth-roll-sd-dist
#| fig-cap: Distribution of the 24-hour rolling standard deviation of depth observations (binwidth = 0.1 m). Dotted orange line indicates $\text{rolling\_sd}_\text{max}$.
#| fig-alt: Histogram of the rolling standard deviation calculated for measured depth observations.

depth_roll_sd <- thresholds %>% 
  filter(variable == "sensor_depth_measured_m", qc_test == "rolling_sd") %>% 
  select(county, threshold, threshold_value) %>% 
  pivot_wider(names_from = threshold, values_from = threshold_value) 

p <- dat_depth %>% 
  filter(!is.na(sd_roll)) %>% 
  plot_histogram(hist_col = sd_roll, binwidth = 0.1, pal = hist_pal) +
  facet_wrap(~county, ncol = 2, scales = "free_y") +
  geom_vline(
    dat = depth_roll_sd, aes(xintercept = rolling_sd_max), 
    linetype = 3, col = "#EDA247"
  ) +
  scale_x_continuous("Measured Sensor Depth (m)\nRolling Standard Deviation") +
  theme_facet_plotly2

ggplotly(p, tooltip = "text")
```

### Rolling SD Thresholds

```{r}
#| label: tbl-depth-sd
#| tbl-cap: Rolling standard deviation threshold for measured depth.

depth_roll_sd %>% 
  datatable(options = dt_options2, rownames = FALSE)
```
:::

## Depth Crosscheck Test

@fig-depth-crosscheck shows the minimum measured depth and the estimated sensor depth at low tide. @fig-depth-crosscheck-hist shows a histogram of the absolute difference between these measurements ($\text{depth\_diff}$).

The distribution of $\text{depth\_diff}$ is skewed right, and so several upper percentile values (90th, 95th, and 99.7th percentile) were evaluated to use as the threshold. The 90th percentile was considered too stringent, while the 99.7th percentile was too lenient. The 95th percentile was considered the most useful threshold.

::: panel-tabset

### Figure 6
```{r}
#| message: false
#| fig-height: 5
#| label: fig-depth-crosscheck
#| fig-cap: The minimum measured depth and the estimated sensor depth at low tide.
#| fig-alt: Scatterplot of the minimum measured depth and estimated sensor depth at low tide for each deployment.

depth_stats <- dat_depth %>% 
  # remove Madeline Point 1 as part of preliminary QC
  # the diff_abs = 22 m
  filter(
    !(station == "Madeline Point 1" &
      deployment_range == "2020-Feb-26 to 2020-Oct-20")
  ) %>% 
  group_by(county, station, deployment_range, sensor_serial_number, depth_log) %>% 
  summarise(min = min(sensor_depth_measured_m)) %>% 
  mutate(
    diff_min = depth_log - min,
    abs_diff_min = abs(diff_min),
    across(where(is.numeric), ~round(.x, digits = 1))) %>% 
  ungroup() %>% 
  mutate(`Depth (m)` = factor(depth_log))

depth_crosscheck <- thresholds %>% 
  filter( qc_test == "depth_crosscheck")

p <- ggplot(
  depth_stats, 
  aes(
    depth_log, min, 
    col = `Depth (m)`,
    text = paste0(
      "depth_log: ",  `Depth (m)`, "\n",
      "min_measured: ", min, "\n",
      "station: ", station, "\n",
      "depl: ", deployment_range))) +
  geom_abline(slope = 1, intercept = 0, col = "grey30") +
  geom_point() +
  scale_y_continuous("Minimum Measured Depth (m)") +
  scale_x_continuous("Estimated Depth at Low Tide (m)") 
  
ggplotly(p, tooltip = "text")
```

### Figure 7
```{r}
#| message: false
#| warning: false
#| fig-height: 5
#| label: fig-depth-crosscheck-hist
#| fig-cap: Distribution of the absolute difference between the minimum measured depth and the estimated sensor depth at low tide (binwidth = 1 m).
#| fig-alt: Histogram of the absolute difference between the minimum measured depth and the estimated sensor depth at low tide (binwidth = 1 m).

p <- plot_histogram(depth_stats, hist_col = abs_diff_min, binwidth = 1) +
   geom_vline(
    xintercept = depth_crosscheck$threshold_value, 
    linetype = 3, col = "#EDA247"
  ) +
  scale_x_continuous("Absolute Difference (m)")

ggplotly(p)
```

### Depth Crosscheck Threshold

```{r}
#| label: tbl-depth-crosscheck
#| tbl-cap: Depth Crosscheck Test threshold for measured depth.

depth_crosscheck %>% 
  select(threshold, threshold_value) %>% 
  pivot_wider(names_from = threshold, values_from = threshold_value) %>% 
  datatable(options = dt_options2, rownames = FALSE)
```
:::






