---
title: "CMAR Measurements"
editor_options: 
  chunk_output_type: console
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi = 200, fig.width = 8)

library(here)

source(here("pages/import_data_and_params.R"))

dat <- dat %>% filter(variable == "Salinity")

st_locations <- st_locations %>% 
  filter(STATION %in% unique(dat$station))
  
```

```{css}
.info.legend.leaflet-control i { 
  float: left; 
  margin-right: 8px;
}
```


This page provides an overview of where and how salinity is measured through CMAR's Coastal Monitoring Program.

Figures and values on this page include data collected up to the end of the 2022 field season. This data underwent preliminary quality control (QC).

## Locations

CMAR has collected salinity data from `r length(na.omit(unique(dat$station)))` stations in `r length(na.omit(unique(dat$county)))` counties (@fig-sal-station-locations). 

The stations in Inverness County are in the Bras d'Or Lake, which is a brackish estuary. The salinity measured at these stations is expected to be substantially different from salinity measured at marine stations.

```{r}
#| label: fig-sal-station-locations
#| fig-height: 7
#| fig-cap: Approximate location of stations with salinity data.
#| fig-alt: Map showing the approximate locations of stations with salinity data, coloured by county.

leaflet(st_locations) %>% 
  addProviderTiles(providers$Esri.OceanBasemap) %>% 
  add_circle_markers_county(county_pal = county_pal, size = 6) %>% 
  add_county_fill_legend(county_pal = county_pal)
```

## Number of Observations
```{r}
n_county_obs <- dat %>% 
  filter(group == "county") %>% 
  group_by(county) %>%  
  summarise(n = sum(n)) %>% 
  mutate(n_percent = round(100 * n / sum(n), digits = 1)) %>% 
  arrange(desc(n_percent))
```

A large proportion of these records are from `r n_county_obs$county[1]` County
(`r n_county_obs$n_percent[1]`%), while a small proportion are from `r n_county_obs$county[nrow(n_county_obs)]` (`r n_county_obs$n_percent[nrow(n_county_obs)]`%) and `r n_county_obs$county[nrow(n_county_obs)-1]` (`r n_county_obs$n_percent[nrow(n_county_obs)-1]`%) counties (@fig-sal-n-obs). 

```{r}
#| label: fig-sal-n-obs
#| fig-height: 8
#| fig-cap: The number of salinity observations in each county.
#| fig-alt: Bar plot showing the number of salinity observations for each county.

p <- dat %>%
  filter(group == "county") %>% 
  plot_n_obs_county()

ggplotly(p)
```

## Depth

The salinity sensors are typically placed in relatively shallow positions on the string (@fig-sal-n-obs-depth). All of the salinity measurements recorded at 10 m were collected at the two Inverness stations for a specific project.

```{r}
#| label: fig-sal-n-obs-depth
#| fig-height: 8
#| fig-cap: Number of salinity observations at each depth (rounded to nearest whole number).
#| fig-alt: Bar plot showing the number of salinity observations at each depth.

p <- dat %>%
  filter(group == "all_depth") %>%
  group_by(depth) %>%
  summarise(n = sum(n)) %>%
  plot_n_obs_depth(facet_county = FALSE)
  
ggplotly(p, tooltip = "text")

```

## Sensors

CMAR uses [aquaMeasure SAL](https://www.innovasea.com/wp-content/uploads/2021/07/Innovasea-Aquaculture-Intelligence-Spec-Sheet-062221.pdf) sensors to measure salinity in practical salinity units (PSU).

These sensors can be susceptible to biofouling, which can cause substantial drift in the measured data over time. Sensor drift is difficult to detect with QC tests [@RN24432], and CMAR does not currently test for sensor drift. Users should evaluate salinity data carefully before analysis. It may be beneficial to compare measurements from consecutive deployments (e.g., the end of one deployment and the beginning of the next) to highlight potential drift in the salinity sensors.

