---
title: "CMAR Measurements"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi = 200, fig.width = 8)

library(here)

source(here("pages/import_data_and_params.R"))

dat <- dat %>% filter(variable == "Temperature")
```

```{css}
.info.legend.leaflet-control i { 
  float: left; 
  margin-right: 8px;
}
```

This page provides an overview of where and how temperature is measured through CMAR's Coastal Monitoring Program.

Figures and values on this page include data collected up to the end of the 2022 field season. This data underwent preliminary quality control (QC).


## Locations

CMAR has collected temperature data from `r length(na.omit(unique(dat$station)))` stations in `r length(na.omit(unique(dat$county)))` counties (@fig-temp-station-locations).

```{r}
#| label: fig-temp-station-locations
#| fig-height: 7
#| fig-cap: Approximate location of stations with temperature data.
#| fig-alt: Map showing the approximate locations of stations with temperature data, coloured by county.

leaflet(st_locations) %>% 
  addProviderTiles(providers$Esri.OceanBasemap) %>% 
  add_circle_markers_county(county_pal = county_pal, size = 6) %>% 
  add_county_fill_legend(county_pal = county_pal)
  
```

## Number of Observations
```{r}
n_county_obs <- dat %>% 
  filter(group == "county") %>% 
  group_by(county) %>%  
  summarise(n = sum(n)) %>% 
  mutate(n_percent = round(100 * n / sum(n), digits = 1)) %>% 
  arrange(desc(n_percent))
```

A large proportion of these records are from `r n_county_obs$county[1]` County
(`r n_county_obs$n_percent[1]` %), while a small proportion are from `r n_county_obs$county[nrow(n_county_obs)]` (`r n_county_obs$n_percent[nrow(n_county_obs)]` %) and `r n_county_obs$county[nrow(n_county_obs)-1]` (`r n_county_obs$n_percent[nrow(n_county_obs)-1]` %) counties (@fig-temp-n-obs). 

```{r}
#| label: fig-temp-n-obs
#| fig-height: 8
#| fig-cap: The number of temperature observations in each county.
#| fig-alt: Bar plot showing the number of temperature observations for each county.


p <- dat %>%
  filter(group == "county") %>% 
  plot_n_obs_county()

ggplotly(p)

```

## Depth

The placement of the temperature sensors depends on the depth of the water at each station. Typically, sensors are fastened 2, 5, 10, 15, and 20 m below the surface at low tide. At deeper stations, additional sensors are placed every 5–10 m. At shallow stations, sensors may be attached nearer to the surface (@fig-temp-n-obs-depth).

Sensors may be placed at other depths for technical reasons or specific research projects (e.g., investigation of the oxycline in Whycocomagh Basin).

```{r}
#| label: fig-temp-n-obs-depth
#| fig-height: 10
#| fig-cap: Number of temperature observations at each depth (rounded to the nearest metre).
#| fig-alt: Bar plot showing the number of temperature observations at each depth.

p <- dat %>%
  filter(group == "all_depth") %>%
  group_by(depth) %>%
  summarise(n = sum(n)) %>%
  plot_n_obs_depth(text_size = 12, facet_county = FALSE)

ggplotly(p, tooltip = "text")
```

## Sensors

CMAR uses several types of sensors to measure temperature (@tbl-temp-sensors):

```{r}
#| label: tbl-temp-sensors
#| tbl-cap: Temperature sensors used in CMAR's Coastal Monitoring Program. Note that some aquaMeasure DOT, and aquaMeasure SAL sensors do not measure pressure.

sensors %>%
  select(-c(sensor_type, sensor_min, sensor_max)) %>% 
  mutate(
    variable = case_when(
      str_detect(variable, "temp") ~ "Temperature",
      str_detect(variable, "dissolved") ~ "Dissolved Oxygen",
      str_detect(variable, "salinity") ~ "Salinity",
      str_detect(variable, "depth") ~ "Pressure (sensor depth)"
    )
  ) %>% 
  pivot_wider(names_from = "variable", values_from = "variable") %>% 
  mutate(
    `Variable(s) Measured` = paste(
      Temperature, `Dissolved Oxygen`, Salinity, `Pressure (sensor depth)`,
      sep = ", "
    ),
    `Variable(s) Measured` = str_remove_all(`Variable(s) Measured`, "NA,|, NA"),
    `Sensor (link to specifications)` = 
      paste0('<a  target=_blank href=', url, '>', sensor_model,'</a>')
  ) %>% 
  select(`Sensor (link to specifications)`, Manufacturer = manufacturer, `Variable(s) Measured`) %>%
  arrange(Manufacturer) %>% 
  datatable(
    options = list(
      dom = 'ft',
      searching = FALSE,
      paging = FALSE,
      pageLength = 500,
      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    ), 
    rownames = FALSE, escape = FALSE
  )
```

<br>

Some of these sensors also measure other variables including dissolved oxygen, salinity, sensor depth, and acoustic detections.

The Vemco VR2AR acoustic receiver is typically the deepest sensor, anchored about 0.5 m above the seafloor. It has an acoustic release that is triggered to retrieve the sensor string. Deployments without VR2AR receivers are usually accessible from the surface, but some are retrieved by dragging or by divers.

Through calibration-validation procedures, CMAR has discovered that VR2AR temperature sensors typically record 0.5-1 °C cooler than other temperature sensors. Since the 2025 field season, a HOBO Pro V2 has been attached to every deployed VR2AR receiver. The VR2AR temperature data is compiled but not published.
